---
title: "Lecture 3: Variable creation, variable attributes, and pipes"
subtitle:  "EDUC 263: Managing and Manipulating Data Using R"
author: Ozan Jaquette
date: 
fontsize: 8pt
classoption: dvipsnames  # for colors
urlcolor: blue
output:
  beamer_presentation:
    keep_tex: true
    toc: true
    slide_level: 3
    theme: default # AnnArbor # push to header?
    #colortheme: "dolphin" # push to header?
    #fonttheme: "structurebold"
    highlight: default # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax highlighting); push to header
    df_print: tibble #default # tibble # push to header?    
    latex_engine: xelatex #  Available engines are pdflatex [default], xelatex, and lualatex; The main reasons you may want to use xelatex or lualatex are: (1) They support Unicode better; (2) It is easier to make use of system fonts.
    includes:
      in_header: ../beamer_header.tex
      #after_body: table-of-contents.txt 
---

<!-- CAN THIS BE MOVED TO SOME KIND OF HEADER FILE? --> 

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction/logistics

### Libraries we will use today

```{r}
library(tidyverse)
```

### Data we will use today

Data on off-campus recruiting events by public universities

```{r}
rm(list = ls()) # remove all objects

#load dataset with one obs per recruiting event
load("../../data/recruiting/recruit_event_somevars.Rdata")

#load dataset with one obs per high school
load("../../data/recruiting/recruit_school_somevars.Rdata")

load("../../data/prospect_list/western_washington_college_board_list.RData")

```
Object \hlgc{df\_event\}

- One observation per university, recruiting event

Object \hlgc{df\_event\}

- One observation per high school (visited and non-visited)

# Factors

## Review data types and structures

### Review data types

Primary __data types__ in R:

- numeric (integer & double)
- character
- logical

R CODE CHUNK WITH EXAMPLES

### Review data structures: vectors

Primary __data structures__ in R are __vectors__ and __lists__

\medskip A __vector__ is a collection of values

- each value in a vector is an __element__
- all elements within vector must have same __data type__

```{r}
a <- c(1,2,3)
a
str(a)
```

You can assign __names__ to elements of a vector, thereby creating a __named vector__

```{r}
b <- c(v1=1,v2=2,v3=3)
b
str(b)
```

### Review data structures: lists

Like vectors, __lists__ are objects that contain __elements__; However, __data type__ can differ across elements within a list; an element of a list can be another list


Examples of lists:
```{r}
list_a <- list(1,2,"apple")
str(list_a)
list_b <- list(1, c("apple", "orange"), list(1, 2, 3))
str(list_b)
```
### Review data structures: lists

Like vectors, elements within a list can be named, thereby creating a __named list__

```{r}
str(list_b) # not named

list_c <- list(v1=1, v2=c("apple", "orange"), v3=list(1, 2, 3))
str(list_c) # named
```

### Review data structures: a data frame is a list

A __data frame__ is a list with the following characteristics:

- All the elements must be __vectors__ with the same __length__
- Data frames are __augmented lists__ because they have additional __attributes__ [described later]

```{r}
list_d <- list(col_a = c(1,2,3), col_b = c(4,5,6), col_c = c(7,8,9))
typeof(list_d)
str(list_d)

df_a <- data.frame(col_a = c(1,2,3), col_b = c(4,5,6), col_c = c(7,8,9))
typeof(df_a)
str(df_a)
```
## Attributes and augmented vectors

### Atomic vectors versus augmented vectors

__Atomic vectors__ [our focus so far]
- [(See figure)](http://r4ds.had.co.nz/diagrams/data-structures-overview.png)
- I think of atomic vectors as "just the data"
- Atomic vectors are the building blocks for augmented vectors

Augmented vectors

- __Augmented vectors__ are atomic vectors with additional __atributes__ attached

__Attributes__

- __Attributes__ are additional "metadata" that can be attached to any object (e.g., vector or list)
- Important attributes in R:
    - __Names__: name the elements of a vector (e.g., variable names)
    - __Class__: How object should be treated by object oriented programming language [discussed below]

Main takaway:

- Augmented vectors are atomic vectors (just the data) with additional attributes attached

### Attributes in vectors

```{r}
vector1 <- c(1,2,3,4)
vector1
attributes(vector1)

vector2 <- c(a = 1, b= 2, c= 3, d = 4)
vector2
attributes(vector2)
```


### Attributes in lists

```{r}
list1 <- list(c(1,2,3), c(4,5,6))
str(list1)
attributes(list1)

list2 <- list(col_a = c(1,2,3), col_b = c(4,5,6))
str(list2)
attributes(list2)

list3 <- data.frame(col_a = c(1,2,3), col_b = c(4,5,6))
str(list3)
attributes(list3)

```
### Object class

```{r}
vector1 <- c(1,2,3,4)
vector1
typeof(vector1)
class(vector1)
attributes(vector1)

vector2 <- c(a = 1, b= 2, c= 3, d = 4)
vector2
attributes(vector2)
typeof(vector2)
class(vector2)


```

# FACTORS

### Factors

__Factors__ are used to display categorical data (e.g., marital status)

- A factor is an __augmented vector__ built by attaching a "levels" attribute to an (atomic) integer vectors

The \hlgc{str()} function is useful for identifying which variables are factors. Let's examine the factor variable \hlgc{ethn\_code}

```{r}
typeof(wwlist$ethn_code)
class(wwlist$ethn_code)
str(wwlist$ethn_code)
```
Note that \hlgc{ethn\_code} has \hlgc{type=integer} and \hlgc{class=factor} because the variable has a "levels" attribute

```{r, results='hide'}
attributes(wwlist$ethn_code)
```

Main takeaway:

- The underlying data are integers but the levels attribute is used to display the data.



### Working with factor variables

```{r, results='hide'}
attributes(wwlist$ethn_code)
```
Refer to categories of a factor by the values of the level attribute rather than the underlying values of the variable

```{r}
count(filter(wwlist,ethn_code==11))

count(filter(wwlist,ethn_code=="White"))
```

If you want to refer to underlying values, then apply \hlgc{as.integer()} function to the factor variable
```{r}
count(filter(wwlist,as.integer(ethn_code)==11))
```
### How to identify the variable values associated with factor levels

MAYBE CUT THIS SLIDE IF YOU CAN'T DO THIS WITHOUT PIPES
```{r}

wwlist %>% count(psat_range) %>% as_factor()
```
```{r}
count(filter(wwlist,as.integer(psat_range)==4))
count(filter(wwlist,psat_range=="1270-1520"))
```


### Some in-class exercise involving factors

```{r}
str(wwlist)
```
### Creating factors [from integer vectors]

Factors are just integer vectors with level attributes attached to them. So, to create a factor:

1. create a vector for the underlying data
1. create a vector that has level attributes
3. Attach levels to the data using the \hlgc{factor()} function

```{r}
a1 <- c(1,1,1,0,1,1,0) #a vector of data
a2 <- c("zero","one") #a vector of labels

#attach labels to values
a3 <- factor(a1, labels = a2)
a3
str(a3)

```

Note: By default, \hlgc{factor()} function attached "zero" to the lowest value of vector \hlgc{a1} because "zero" was the first element of vector \hlgc{a2}

### Creating factors [from integer vectors]

Let's turn an integer variable into a factor variable in the \hlgc{wwlist} data frame

Create integer version of \hlgc{sex}
```{r}
wwlist$sex_int <- as.integer(wwlist$sex)
str(wwlist$sex_int)
#wwlist %>% count(sex) %>% as_factor()
```
Assume we know that 1=female, 2=male, 3=unknown

Assign levels to values of integer variable
```{r}
wwlist$sex_int <- factor(wwlist$sex_int, labels=c("female","male","unknown"))
str(wwlist$sex_int)
str(wwlist$sex)
```

### Create factors [from string variables]

To create a factor variable from string variable

1. create a character vector containing underlying data
1. create a vector containing valid levels
3. Attach levels to the data using the \hlgc{factor()} function

```{r}
#underlying data: months my fam is born
x1 <- c("Jan", "Aug", "Apr", "Mar")
#create vector with valid levels
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
#attach levels to data
x2 <- factor(x1, levels = month_levels)
```
Note how attributes differ
```{r}
str(x1)
str(x2)
```
Sorting differs
```{r}
sort(x1)
sort(x2)
```

### Create factors [from string variables]

Let's create a character version of variable \hlgc{sex} and then turn it into a factor

```{r}
#Create character version of sex
wwlist$sex_char <- as.character(wwlist$sex)

#investigate character variable
str(wwlist$sex_char)
table(wwlist$sex_char)

#create new variable that assigns levels
sex_fac <- factor(wwlist$sex_char, levels = c("F","M","U"))
str(wwlist$sex_char)
```
How the \hlgc{levels} argument works when underlying data is character

- Matches value of underlying data to value of the level attribute
- Converts underlying data to integer, with level attribute attached

\medskip See chapter 15 of Wickham for more on factors (e.g., modifying factor order, modifying factor levels)

### Substantial exercise on using/creating factors, using either df_school or df_event datasets


# Labeling variables


# Pipes

### What are "pipes", %>%

__Pipes__ are a means of perfoming multiple steps in a single line of code

- Pipes are part of __tidyverse__ suite of packages, not __base R__
- When writing code, the pipe symbol is \hlgc{\%>\%}
- Basic flow of using pipes in code:
    - \hlgc{object \%>\% some\_function \%>\% some\_function, \ldots}    
- Pipes work from left to right:
    - The object/result from left of \hlgc{\%>\%} pipe symbol is the input of function to the right of the \hlgc{\%>\%} pipe symbol 
    - In turn, the resulting output becomes the input of the function to the right of the next \hlgc{\%>\%} pipe symbol 

### Do some tasks with and without pipes

Print data for "first-generation" prospects
```{r, results='hide'}
filter(wwlist, firstgen == "Y")
wwlist %>% filter(firstgen == "Y")
```
Comparing the two approaches:

- In the "without pipes" approach, the object is the first argument \hlgc{filter()} function
- In the "pipes" approach, you don't specify the object as the first argument of \hlgc{filter()}
    - Why? Because \hlgc{\%>\%} "pipes" the object to the left of the \hlgc{\%>\%} operator into the function to the right of the \hlgc{\%>\%} operator
    
Main takeaway:

- Whenever you write code using pipes, functions to the right of a \hlgc{\%>\%} pipe operator should not explicitly name the object that is the input to that function. Rather, the object to the left of the \hlgc{\%>\%} pipe operator is automatically the input.
    
### Do some tasks with and without pipes

Print data for "first-generation" prospects for selected variables [output omitted]

```{r, results='hide'}
select(filter(wwlist, firstgen == "Y"), state, hs_city, ethn_code)

wwlist %>% filter(firstgen == "Y") %>% select(state, hs_city, ethn_code)
```
Comparing the two approaches:

- In the "without pipes" approach, code is written "inside out"
    - The first step in the task -- identifying the object -- is the innermost part of code
    - The last step in task -- selecting variables to print -- is the outermost part of code
- In "pipes" approach the left-to-right order of code matches how we think about the task
    - First, we start with an object __*and then*__ (\hlgc{\%>\%}) we use \hlgc{filter()} to isolate first-gen students __*and then*__ (\hlgc{\%>\%}) we select which  variables to print

### Do some tasks with and without pipes

Count the number "first-generation" prospects from the state of Washington

```{r}
count(filter(wwlist, firstgen == "Y", state == "WA"))

wwlist %>% filter(firstgen == "Y", state == "WA") %>% count()
```

### Do some tasks with and without pipes [last example]

Create frequency table of \hlgc{sex} for "first-generation" prospects from WA
```{r}
wwlist_temp <- filter(wwlist, firstgen == "Y", state == "WA")
table(wwlist_temp$sex, useNA = "ifany")

wwlist %>% filter(firstgen == "Y", state == "WA") %>% count(sex)
```
Comparison of two approaches

- without pipes, task requires multiple lines of code; this is quite common
    - first line creates object; second line analyzes object
- with pipes, task can be completed in one line of code and you aren't left with objects you don't care about

Note: the pipes approach above is a useful way to show the __values__ associated with each __factor level__ for factor variables

### Student exercises with pipes

CREATE STUDENT EXERCISES

# Creating variables using mutate

### Our plan for learning how to create new variables

Recall that the \hlgc{dplyr} package within the \hlgc{tidyverse} provide a set of functions that can be described as "verbs":

- __subsetting__, __sorting__, and __transforming__

What we've done | Where we're going
------------- | -------------
__Subsetting data__ | __Transforming data__
- \hlgc{select()} variables  | - \hlgc{mutate()} creates new variables
- \hlgc{filter()} observations | - \hlgc{summarize()} calculates across rows
__Sorting data__ | - \hlgc{group\_by()} to calculate across rows within groups
- \hlgc{arrange()} | 

__Today__

- we'll use \hlgc{mutate()} to create new variables based on calculations across columns within a row

__Next week__

- we'll combine \hlgc{mutate()} with \hlgc{summarize()} and \hlgc{group\_by()} to create variables based on calculations across rows

### STUFF TO ADD TO MUTATE

- take some time to go through syntax of help file
- ifelse() function
- recode() function
- case_when() function
- go through using recode() function with mutate?

### Introduce \hlgc{mutate()} function

\hlgc{mutate()} creates new columns (variblaes) that are functions of existing columns

Description of \hlgc{mutate()}

- Every row of data is retained
- Is the __tidyverse__ approach to creating variables, not the __Base R__ approach
- Works best with pipes \hlgc{\%>\%}

We'll create variables from data frame \hlgc{df\_school} which has one observation for each high school

- Task: create pct of students on free/reduced lunch (output omitted)

```{r, results='hide'}
school_sml <- df_school %>% filter(school_type == "public") %>% 
    select(ncessch, num_fr_lunch, total_students)

school_sml %>% mutate(pct_fr_lunch = num_fr_lunch/total_students)
```

Can combine \hlgc{(select())} with pipes \hlgc{\%>\%} to control which columns printed, so no need to create dataset with fewer columns. But let's create data frame with public high schools only

```{r}
school_pub <- df_school %>% filter(school_type == "public")
```

### Introduce \hlgc{mutate()} function

New variable not retained unless we __assign__ \hlgc{<-} it to an object (existing or new)

```{r, results='hide'}
school_pub %>% mutate(pct_fr_lunch = num_fr_lunch/total_students)
names(school_pub)

school_pub_temp <- school_pub %>% 
  mutate(pct_fr_lunch = num_fr_lunch/total_students) 
names(school_pub_temp)
```

How to create percent free/reduced lunch in __Base R__

```{r}
school_pub$pct_fr_lunch <- school_pub$num_fr_lunch/school_pub$total_students
```
\hlgc{mutate()} can create multiple variables at once

```{r, results='hide'}
school_pub %>% 
  mutate(pct_fr_lunch = num_fr_lunch/total_students,
         pct_prof_math= num_prof_math/num_took_math) %>%
  select(num_fr_lunch, total_students, pct_fr_lunch, 
         num_prof_math, num_took_math, pct_prof_math)
```

### Student exercise using mutate()


### Mutate to create indicator variables

We often create dichotomous (0/1) indicator variables of whether something happened (or whether something is TRUE)

- Variables that are of substantive interest to project
    - e.g., did student graduate from college
- Variables that help you investigate data, check quality
    - e.g., indicator of whether an observation is missing/non-missing for a particular variable

Let's conduct some investigations of \hlgc{df\_school}, which has one observation for each high school

Rename some variables (output omitted)
```{r, results='hide'}
str(df_school)
df_schoolv2 <- df_school %>% 
  rename(
    visits_berkeley = visits_by_110635,
    visits_boulder = visits_by_126614,
    visits_bama = visits_by_100751,
    state_berkeley = inst_110635,
    state_boulder = inst_126614,
    state_bama = inst_100751)

names(df_schoolv2)
```


### Creating indicators for \hlgc{df\_schoolv2} data frame

Create TRUE/FALSE indicator that median household income greater than $50,000
```{r}
df_schoolv2_temp <- df_schoolv2 %>% mutate(incgt50k = avgmedian_inc_2564>50000)

df_schoolv2_temp %>% select(avgmedian_inc_2564, incgt50k) %>% head(n=3)

df_schoolv2_temp %>% filter(is.na(avgmedian_inc_2564)) %>% count(incgt50k)
```
Important takeaway:

- Variable created by \hlgc{mutate()} equals \hlgc{NA} for obs if input variable to \hlgc{mutate()} is missing for that obs. This is a good thing!

### Creating indicators for \hlgc{df\_schoolv2} data frame

Create TRUE/FALSE indicator that school is less than 50 percent white

```{r}
df_schoolv2_temp <- df_schoolv2 %>% mutate(lt50pctwhite = pct_white<50)
df_schoolv2_temp %>% select(pct_white,lt50pctwhite) %>% head(n=3)
str(df_schoolv2_temp$lt50pctwhite)
```
Create 0/1 integer indicator rather than logical indicator 
```{r}
df_schoolv2_temp <- df_schoolv2 %>% mutate(lt50pctwhite = as.integer(pct_white<50))
df_schoolv2_temp %>% select(pct_white,lt50pctwhite) %>% head(n=3)
str(df_schoolv2_temp$lt50pctwhite)
```

### Student exercises

0/1 indicators of whether school received visit from each university
```{r}
df_schoolv2 %>% count(visits_berkeley)
df_schoolv2_temp <- df_schoolv2 %>% mutate(yesvis_berkeley = visits_berkeley>0)

df_schoolv2_temp %>% filter(visits_berkeley>0) %>% select(visits_berkeley,yesvis_berkeley)
df_schoolv2_temp %>% filter(visits_berkeley==0) %>% select(visits_berkeley,yesvis_berkeley)

```


### Investigating \hlgc{wwlist} data frame

?MAYBE CUT UNTIL NEXT LECTURE AND HAVE SECTION ON USING DESCRIPTIVE STATS TO INVESTIGATE DATA?

\medskip Let's conduct some investigations of \hlgc{wwlist}, which is frankly a pretty weird dataset!

- When conducting investigations, really important to be careful about missing values

```{r, results='hide'}
str(wwlist)
```

Variable \hlgc{receive\_date} indicates date prospect list data received from College Board

```{r}
wwlist %>% count(receive_date)
```


### other 

```{r}
#num_took_math, num_prof_math
ww_narrow <- wwlist %>% select(psat_range, contains("gpa"), state, zip, for_country, ethn_code, homeschool, firstgen)
load("../../data/recruiting/recruit_event_somevars.Rdata")
```

