---
urlcolor: blue
output:
  pdf_document:
    keep_tex: true
    fig_caption: yes
    latex_engine: pdflatex
    template: svm-r-markdown-template_ua.tex
geometry: margin=1in

title: "HED 696C: Data Management and Manipulation in R"
author: The University of Arizona
date: "Fall 2019"

instructor: "Karina Salazar"
pronouns: "she/her/hers"
email: "ksalazar@email.arizona.edu"
officehours: Tues 3-4PM; and by appt
office: "EDU 313"

web: "ozanj.github.io/rclass/"
discuss: "d2l.arizona.edu"
classroom: "COE 331"
classhours: "Wednesdays 4:15 - 6:45 pm"

ta: "Patricia Martín"
tapronouns: "she/her/hers"
taemail: "pmarti@g.ucla.edu"
taofficehours: Wed 12-1PM; Thur 1:30-2:30PM
location: "Moore Hall 3120 (computer lab)"

  
fontfamily: mathpazo
fontsize: 11pt
header-includes:
   - \linespread{1.05}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
#template: ~/Desktop/GitHub/rclass/syllabus/svm-r-markdown-template.tex
```

```{r, eval=FALSE}
fri <- as.Date("2018-09-24")


advdate <- function(obj, adv) {
 tmon <- obj + 7*(adv-1)
 tfri <- obj + 4 + 7*(adv-0)
 tmon <- format(tmon, format="%m/%d")
 tfri <- format(tfri, format="%m/%d")
 zadv <- sprintf("%02d", adv)
 if(adv == 11) {
   tmp <- paste("Finals Week",sep='',", ", tmon, "-" ,tfri)
 } else {
   tmp <- paste("Week ",zadv,sep='',", ", tfri)
 }
 return(tmp)
}
advdate(fri, 11)
```


```{r}
fri <- as.Date("2018-09-24")

advdate <- function(obj, adv) {
 tmon <- obj + 7*(adv-1)
 tfri <- obj + 4 + 7*(adv-0)
 tmon <- format(tmon, format="%m/%d")
 tfri <- format(tfri, format="%m/%d")
 zadv <- sprintf("%02d", adv)
 #print(zadv)
 #print(str(zadv))
 if(adv == 11) {
   tmp <- paste("Finals Week",sep='',", ",tfri)
 } else if (adv==8) {
   tmp <- paste("Thanksgiving",sep='',", ", tfri)
   
 } else if (adv<8) {
   lec=adv+1
   tmp <- paste("Lecture ",lec,sep='',", ", tfri)
 } else {
   tmp <- paste("Lecture ",adv,sep='',", ", tfri)
 }
 return(tmp)
}
#advdate(fri, 11)
```
# Course Description

This course has two foundational goals: (1) to develop core skills in “data management,” which are important regardless of which programming language you use, and (2) to learn the fundamentals of the R programming language.

Data management consists of acquiring, investigating, cleaning, combining, and manipulating data.  Most statistics courses teach you how to analyze data that are ready for analysis. In real research projects, cleaning the data and creating analysis datasets is often more time consuming than conducting analyses.  This course teaches the fundamental data management and data manipulation skills necessary for creating analysis datasets.

The course will be taught in R, a free, open-source programming language. R has become the most popular language for statistical analysis, surpassing SPSS, Stata, and SAS. What differentiates R from these other languages is the thousands of open-source “libraries” created by R users.  R is one of the most popular languages for “data science,” because R libraries have been created for web-scraping, mapping, network analysis, etc.  By learning R you can be confident that you know a programming language that can run any modeling technique you might need and has amazing capabilities for data collection and data visualization. By learning fundamentals of R in this course, you will be "one step away" from web-scraping, network analysis, interactive maps, quantitative text analysis, or whatever other data science application you are interested in.

Students will become proficient in data manipulation tasks through weekly “problem sets” that you complete in groups of three. These problem sets will account for 80% of your grade for the course.  Each week class will begin with one group will leading a discussion of challenges they encountered while completing the problem set.  The rest of class time will be devoted to learning new material. The instructor will provide students with lecture notes, and also data and code used during lecture. Therefore, student can follow along by running code from their own computers.

# Course Learning Goals

1. Understand fundamental concepts of object oriented programming
    - What are the basic object types and how do they apply to statistical analysis
    - What are object attributes and how do they apply to statistical analysis
1. Become familiar with Base R approach to data manipulation and Tidyverse approach to data manipulation
1. Investigate data patterns 
    -	Sort datasets in ways that generate insights about data structure
    - Select specific observations and specific variables in order  to identify data structure and to examine whether variables are created correctly
    -	Create summary statistics of particular variables to diagnose errors in data
1. Create variables
    - Create variables that require calculations across columns
    - Create variables that require processing across rows
1. Combine multiple datasets
    - Join (merge) datasets
    - Append (stack) datasets
1. Manipulate the organizational structure of datasets
    - summarize and collapse by group
    - Tidy untidy data
1. Automate iterative tasks
    - Write your own functions
    - Write loops
1. Learn habits of mind and practical strategies for cleaning dirty data and avoiding errors when creating analysis variables variables

# Prerequisite Requirements 

1.	Students must have taken at least a one-semester introductory statistics course.  

1.	Students should have some very basic experience using statistical programming software (e.g., SPSS, Stata, R, SAS). 

1.	[General computer skills] Students should be able to download files from the internet, rename these files, save them to a folder of your choosing, and open this folder.  
    - During this course we will often be downloading datasets, opening .Rmd files and .R scripts, changing directories to the folder where we stored the data, and then opening the dataset we just downloaded.  Therefore, it is important that students feel comfortable doing these tasks. 


# Course Readings

Course readings will be assigned from:

- Wickham, H., & Grolemund, G. (2018). _R for Data Science_. Retrieved from http://r4ds.had.co.nz/ [FREE!]
- Xie, Y., Allaire, J. j., & Grolemund, G. (2018). _R Markdown: The Definitive Guide_.  Retrieved from https://bookdown.org/yihui/rmarkdown/ [FREE!]

# Required Software and Hardware

## Software [FREE!]

Instructions on downloading software can be found [here](https://github.com/ozanj/rclass/raw/master/educ263_todo.pdf)

Please install the following software on your laptop

- R
- RStudio
- MikTeK/MacTeX

## Hardware

- Please bring in laptop with above software installed each week

# Course Website and Resources

Course Website can be found [here](https://ozanj.github.io/rclass/). We wil use this website to download course materials such as lecture slides in pdf and .Rmd formats, data, weekly problem sets, and other class resources.  


#Discussion and Homework Questions

We are using D2L as our class discussion forum where folks can ask homework questions/comments to share with the instructor and the entire class. If you're stuck on a homework question or are experiencing problems with R more generally odds are others are too. Posting questions and concerns on D2L is the easiest way for us to all benefit from each other's knowledge. When asking questions on D2L, please include as many details to replicate the "error." Always indicate the homework assignment and question number that's causing you issues, insert code, screenshots, and text to your posts. 

I strongly encourage all questions related to course content to be posted on the D2L discussion forum for each week. I will do my best to reply to all posts within 24 hours. I also encourage you all to share your thoughts/answers on posts by your classmates. Writing out explanations to student questions will improve your own knowledge and will benefit your classmates. Sharing different ways to get at the "right" answer will be beneficial for all. 

# Assignments & Grading  

Your final grade will be based on the following components:

-	Weekly problem sets (90 percent of total grade)
-	Attendance and participation (10 percent of total grade)

## Weekly problem sets

Problem sets are due by 4:15PM each Wednesday (right before the class meeting).  Late submissions will not receive points because we will discuss solutions during class.  The lowest grade will be dropped from the calculation of your final grade.  

In general, each problem set will give you practice using the skills and concepts introduced during the previous lecture.  For example, after the lecture on joining (merging) datasets, the problem set for that week will require that students complete several different tasks involving merging data.  Additionally, the weekly problem sets will require you to use data manipulation skills you learned in previous weeks.  

Students can work on problem sets with other students.  However, each student will submit their own assignment.  You are encouraged to share ideas and get help from your classmates However, it is important that you understand how to do the problem set on your own, rather than copying the solution developed by group members.

A general strategy I recommend for completing the problem sets is as follows: (1) after lecture, do the reading associated with that lecture; (2) try doing the problem set on your own; (3) talk/meet with classmates to work through the problem set, with a particular focus on areas group members find challenging.  


## Attendance and Participation (10 percent of total grade)

Students are required to attend the weekly class meetings.  Each week will cover material and skills that will be used and necessary for tasks throughout the semester.  If you need to miss a class meeting for any professional or personal reason, please let me know beforehand (if possible). Excused absences for professional or personal reasons will not result in a loss of attendance points.  However, you will be responsible for all material covered in that class and you will be expected to turn in homework assignments on time. 

Students are expected to participate in the weekly class meetings by being attentive, by asking questions, by answering questions posed by classmates and by the professor. In addition to participation during class meetings, students can receive strong participation grades by asking questions and answering questions on D2L. 

# Course Policies  

## Classroom environment

We all have a responsibility to ensure that every member of the class feels valued, safe, and included. 

With respect to the course material, learning programming and the essential skills of data manipulation is hard! This stuff feels overwhelming to me all the time. So it is important that we all create an environment where students feel comfortable asking questions and talking about what they did not understand. 

With respect to creating an inclusive environment, be mindful that what you say affects other people. So express your thoughts in a way that doesn’t make people feel excluded.


## Online Collaboration/Netiquette

You will communicate with instructors and peers virtually through a variety of tools such as discussion forums, email, and web conferencing. The following guidelines will enable everyone in the course to participate and collaborate in a productive, safe environment.

-	Be professional, courteous, and respectful as you would in a physical classroom.
-	Online communication lacks the nonverbal cues that provide much of the meaning and nuances in face-to-face conversations. Choose your words carefully, phrase your sentences clearly, and stay on topic.


## Accessibility and Accommodations 

At the University of Arizona, we strive to make learning experiences as accessible as possible. If you anticipate or experience barriers based on disability or pregnancy, please contact the Disability Resource Center (520-621-3268, https://drc.arizona.edu/) to establish reasonable accommodations.

## Academic Honesty: 

Academic Integrity at the University of Arizona is the principle that stands for honesty and ethical behavior in all homework, tests, and assignments.  All students should act with personal integrity and help to create an environment in which all can succeed.

Violations of the UA Code of Academic Integrity are serious offenses.  As your instructor, I will deal with alleged violations in a fair and honest manner.  As students, you are expected to do your own work and follow class rules on all tests and assignments unless I indicate differently.  Alleged violations of the UA Code of Academic Integrity will be reported to the Dean of Students Office and will result in a sanction(s) (i.e., loss of credit on assignment, failure in class, suspension, etc.)  

Students should review the UA Code of Academic Integrity which can be found at: https://deanofstudents.arizona.edu/policies/code-academic-integrity


# Course Schedule and Required Reading
In the below schedule, I lecture on a topic, and then you do the reading about that topic and are required to complete a problem set about that topic. However, if you would prefer to the reading about a topic __prior__ to me lecturing about that topic, feel free to do so.



## Lecture 1, 8/28/2019: Course introduction; objects in R

- Reading (after class): Wickham and Grolemund (W&G) 1; W&G 2; W&G 4; W&G 20.1 - 20.3
- RmD 

## `r advdate(fri, 1)`: Investigating data patterns

- Problem set due (before class): Yes
- Reading (after class): 
    - GW 5.1 - 5.4
    - Xie, Allaire & Grolemund (XAG) 3.1 [LINK HERE](https://bookdown.org/yihui/rmarkdown/html-document.html)
    - Spend 15 minutes studying the "R Markdown Reference Guide" [LINK HERE](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

## `r advdate(fri, 2)`: Pipes and variable creation

- Problem set due (before class): Yes
- Reading (after class): 
    - GW 5.5 (creating variables)
    - XAG 3.3 (R Markdown, creating PDF documents) [LINK HERE](https://bookdown.org/yihui/rmarkdown/pdf-document.html)
        - note: sections 3.3.5 through 3.3.8 will feel somewhat cryptic and are not required for this course; so just do the best you can with those

## `r advdate(fri, 3)`: Processing across rows

- Problem set due (before class): Yes
- Reading (after class): 
    - GW 5.6 - 5.7 (grouped summaries and mutates)
    - XAG 4.1 (R Markdown, ioslides presentations) [LINK HERE](https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html) and 4.3 (R Markdown, Beamer presentations) [LINK HERE](https://bookdown.org/yihui/rmarkdown/beamer-presentation.html)

## `r advdate(fri, 4)`: Augmented vectors, Survey data, and exploratory data analysis

- Problem set due (before class): Yes
- Reading (after class): 
    - GW 15.1 - 15.2 (factors) [this is like 2-3 pages]
    - [OPTIONAL] GW 15.3 - 15.5 (remainder of "factors" chapter)
    - [OPTIONAL] GW 20.6 - 20.7 (attributes and augmented vectors)
    - [OPTIONAL] GW 10 (tibbles)

## `r advdate(fri, 5)`: Tidy data

- Problem set due (before class): Yes
- Reading (after class):
    - Work through slides we from lecture 6 that we don't get to in class
        - [REQUIRED] slides from section 4 "Tidying data"
        - [OPTIONAL] slides from section 5 "Missing data"
    - [OPTIONAL] GW chapter 12 (tidy data)
        - Lecture 6 covers this material pretty closely, so read chapter if you can, but I get it if you don't have time
    - [OPTIONAL] Wickham, H. (2014). Tidy Data. _Journal of Statistical Software_, 59(10), 1-23. doi:10.18637/jss.v059.i10
        - This is the journal article that introduced the data concepts covered in GW chapter 12 and created the packages related to tidying data
        - Link to article here: [LINK](https://www.jstatsoft.org/article/view/v059i10)

## `r advdate(fri, 6)`: Joining multiple datasets

- Problem set due (before class): Yes
- Reading (after class): GW 13

## `r advdate(fri, 7)`: Acquiring data

- Problem set due (before class): Yes
- Reading (after class): GW 11

## `r advdate(fri, 8)`: No class

## `r advdate(fri, 9)`: Writing functions

- Problem set due (before class): Yes
- Reading (after class): GW 19

## `r advdate(fri, 10)`: Accessing object elements and looping

- Problem set due (before class): Yes
- Reading (after class): GW 20.4 - 20.5; 21.1 - 21.3

## `r advdate(fri, 11)`: No class

- Problem set due: Yes

# Campus Resources  
**Counseling and Psychological Services (CAPS)**  
This is a multidisciplinary student mental health center for the UCLA campus. CAPS offers an array of free services including individual counseling. If you suspect you are experiencing mental health problems or just need someone to talk to, you can make an appointment at John Wooden Center West, facing Drake Stadium, second floor, 310-825-0768. http://www.counseling.ucla.edu

**Report Discrimination**
UCLA is committed to maintaining a campus community that provides the stronget possible support for the intellectual and personal growth of all its members- students, faculty, and staff. Acts intended to create a hostile climate are unacceptable. To file an online incident report, visit: https://equity.ucla.edu/report-an-incident/  

**Sexual harassment**  
Title IX prohibits gender discrimination, including sexual harassment, domestic and dating violence, sexual assault, and stalking. If you have experienced sexual harassment or sexual violence, you can receive confidential support and advocacy at the CARE Advocacy Office for Sexual and Gender-Based Violence. 1st Floor Wooden Center West, via email CARE.advocate@careprogram.ucla.edu or by phone (310) 206-2465. In addition, Counseling and Psychological Services (CAPS) provides confidential counseling to all students and can be reached 24/7 at (310) 825-0768. You can also report sexual violence or sexual harassment directly to the University's Title IX Coordinator, 2241 Murphy Hall, via email at mcato@equity.ucla.edu, or via phone at (310) 206-3417. Reports to law enforcement can be made to UCPD at (310) 825-1491. *Faculty and TAs are required under the UC Policy on Sexual Violence and Sexual Harassment to inform the Title IX Coordinator should they become aware that you or any other student has experienced sexual violence or sexual harassment.*  

**LGTBQ Resource Center**
This resource center provides a range of education and advocacy services supporting intersectional identity development. It fosters unity; wellness; and an open, safe, inclusive environment for lesbian, gay, bisexual, intersex, transgender, queer, asexual, questioning, and same-gender-loving students, their families, and the entire campus community. Find it in the Student Activities Center, or via email lgbt@lgbt.ucla.edu. Visit their website for more information: https://www.lgbt.ucla.edu/  

**International Students**  
The Dashew Center provides a range of programs to promote cross-cultural learning, language improvement, and cultural adjustment. Their programs include trips in the LA area, performances, and on-campus events and workshops. Visit their website for more information: https://www.internationalcenter.ucla.edu/  

**Undocumented Student Program**  
This program provides a safe space for undergraduate and graduate undocument students. USP supports the UndocuBruin community through personalized services and resources, programs, and workshops. Visit their website for more information: https://www.usp.ucla.edu/  

**Student legal services**  
UCLA student legal services provides a range of legal support to all registered and enrolled UCLA students. Some of their services include:  

- Landlord/Tenant Relations  
- Accident and Injury Problems  
- Domestic Violence and Harassment  
- Divorces and Other Family Law Matter  

For more information visit their website: http://www.studentlegal.ucla.edu/index.php

**Students with dependents**  
UCLA Students with Dependents provides support to UCLA studens who are parents, guardians, and caregivers. Some of their services include:  

- Information, referrals, and support to navigate UCLA (childcare, family housing, financial aid)  
- Access to information about resources within the larger community  
- On-site application and verification for CalFresh (food stamps) & MediCal and assistance with Cal
Works/GAIN  
- A quiet study space  
- Family friendly graduation celebration in June  

For more information visit their website: https://www.deanofstudents.ucla.edu/Portals/16/Documents/studentsdependents.pdf  

**Lactation Rooms**  
[Map to lactation rooms on campus](https://map.ucla.edu/?f=5)  

**Gender Inclusive restrooms**  
[Map to gender inclusive restrooms](http://maps.ucla.edu/campus/?a_layers=flyr_restrooms)  

**Campus accessibility**  
[Campus accessibility map](https://maps.ucla.edu/downloads/pdf/Access_08_21_15.pdf)



