---
title: "Lecture 2 problem set"
author: "INSERT YOUR NAME HERE"
date: "9/28/2018"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```
# Make changes to YAML header

Recommend reading R Markdown: The Definative Guide section 3.1 ([LINK HERE](https://bookdown.org/yihui/rmarkdown/html-document.html)) before answering these questions

1. Add a table of contents to YAML header
1. table of contents should have "depth" of 2
1. table of contents should "float"
1. add section numbering to headers


Load tidyverse package
```{r}
#install.packages("tidyverse") #install if you do not have tidyverse installed
library(tidyverse)
```

Load data frame objects
```{r}
rm(list = ls()) # remove all objects
getwd()

load(url("https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_school_allvars.RData"))
#glimpse(df_school_all)

df_school_all %>% count(total_visits)
```


# Investigating objects, Base R approach

1. Answer the following questions about the object `df_school_all` by running the appropriate R command within text (as opposed to in an R code chunk) and writing any substantive response required by the question. The first question will be answered for you to show how it works
    + What "type" of object is `df_school_all`?
        + ANSWER [GIVEN]: the object `df_school_all` has type equals `r typeof(df_school_all)`
        type
    + What is the "length" of the object `df_school_all`? What does this specific value of length refer to? (Hint: Use `r ` to run code within text)
        + ANSWER 
    + how many "rows" are in the object `df_school_all`? what does each row represent?  
        + ANSWER 
        
        
\newline  

2. In the below code chunk, use the `str()` function to describe the contents of `df_school_all` and then answer the following questions in text above the code chunk.  
    + What does each element of the object `df_school_all` represent?  
    
    + Are the individual elements within `df_school_all` lists or vectors?  
    
    + Are the individual elements within `df_school_all` named or un-named? what do these element names refer to?  
    
    
```{r}

```

\newline

3. These questions refer to the variable `school_type` within the object `df_school_all`. For the first two questions, run the appropriate R command within text (as opposed to in an R code chunk) and write any substantive response required by the question
    + What is the data "type" of `school_type`?
        + ANSWER
    + What is the "length" of `school_type`? What does this specific value of length refer to?
        + ANSWER
        
4. In these questions, you will apply the `table()` function to the variable `school_type` within the object `df_school_all`.  
    + In your own words, what does the `table()` function do?  
        + ANSWER
    + What does the `useNA` argument of the `table()` function control?
        + ANSWER
    + What is the default value of the `useNA` argument and what does this default value mean?
        + ANSWER
    + What happens when you assign the value `"ifany"` to the `useNA` argument?
        + ANSWER
    + What happens when you assign the value `"always"` to the `useNA` argument?
        + ANSWER
    + In the below R code chunk, use the `table()` function to count the number of observations for each value `school_type` three different ways: 
        + First, without specifying any value for `useNA`
        + Second, by assigning the value `"ifany"` to the `useNA` argument
        + Third,  by assigning the value `"always"` to the `useNA` argument
```{r}

```

# Filter, select, arrange questions

The data frame `df_school_all` has one observation for each high school (public and private). 

- The variables that begin with `visits_by_...` identify how many off-campus recruiting visits the high school received from a particular public university. For example, UC Berkeley has the ID `110635` so the variable `visits_by_110635` identifies how many visits the high school received from UC Berkeley. 
- The variable `total_visits` identifies the number of visits the high school received from all (16) public research universities in this data collection sample.

For the questions below, imagine that you have been asked by a major news outlet to identify which high schools receive the most total number of off-campus recruiting visits from public universities. 

- for all questions below, you can answer using one line of code or you can answer in several steps (e.g., first create new data frame, then print selected variables)
- For questions that ask you to print the "top 10" observations, you can simply print the object and rely on the fact that the default option [for "tibble" data frames] is to print 10 observations OR you can wrap the command in the `head()` function and explicitly tell R to print 10 observations
- Before conducting analyses, we'll rename the variable `avgmedian_inc_2564` to give it a shorter name.

1. Rename the variable `avgmedian_inc_2564` to `med_inc` and assign new variable name to the existing object `df_school_all`
```{r}

```



2. The news outlet is interested in comparing the in-state and out-of-state high school visits for [The University of Alabama](https://nces.ed.gov/globallocator/col_info_popup.asp?ID=100751) (IPEDS ID = 100751) variable `visits_by_100751`. Count the number of in-state public high school visits for The University of Alabama.   
    - Note: You will need to use `filter` and the `count` function  
    - Use commas to separate varibles for this questions (e.g. filter(dataframe, variable == something, variable == something))
    - You can do this in one step by wrapping the `count` function around the `filter` function ; or you can do this in two steps by creating a new data frame first  
```{r}

```




3. How many public in-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?
```{r}

```



4. Now count the out-of-state public high school visits for the University of Alabama without using commas.  
    - hint: `&` or `%in%`
```{r}

```




5. How many public out-of-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?
```{r}

```




6. The news outlet is interested in getting more information about the high schools these public research universities visit. The variable `total_visits` counts the total number of visits to the high school across all public 16 public research universities in the sample. Print the `name`, `state_code`, `city`, `school_type`,`total_visits`, `med_inc`, `pct_white`, `pct_black`, `pct_hispanic`, `pct_asian`, `pct_amerindian`, and `pct_other` of the top 10 schools with respect to the variable `total_visits` (top 10 schools visited)
    - Note: You will have to sort descending by total number of visits
    - You can do this in one step by wrapping the `select()` function around the `arrange()` (i.e., sort) function; or you can do this in two steps by creating a new data frame first
```{r}

```




7. Print the following (same vars as above):
    - (A) the top 10 public high schools in terms of total number of visits and then
    - (B) the top 10 private high schoools in terms of total number of visits
    
```{r}

```



8. Print the following (same vars as above): 
    - (A) the top 10 public high schools in Massachusetts in terms of total number of visits and then 
    - (B) the top 10 private high schools in Massachusetts in terms of total number of visits
    - (C) the top 10 public high schools in California in terms of total number of visits and then 
    - (D) the top 10 private high schools in California in terms of total number of visits    

```{r}


```
  

Once finished, knit to (HTML) and upload both .Rmd and HTML files to class website under the week 2 tab  
*Remember to use this naming convention "lastname_firstname_ps2"*





