---
title: "Lecture 2 problem set"
author: "INSERT YOUR NAME HERE"
date: "9/28/2018"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```
# Make changes to YAML header

Recommend reading R Markdown: The Definative Guide section 3.1 ([LINK HERE](https://bookdown.org/yihui/rmarkdown/html-document.html)) before answering these questions

1. Add a table of contents to YAML header
1. table of contents should have "depth" of 2
1. table of contents should "float"
1. add section numbering to headers  

Note: you can knit this document to HTML at any point to give you an idea of how the compiled HTML document is displayed. 


Load tidyverse package
```{r}
#install.packages("tidyverse") #install if you do not have tidyverse installed
library(tidyverse)
```

Load data frame objects
```{r}
rm(list = ls()) # remove all objects
getwd()

load(url("https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_school_allvars.RData"))
#glimpse(df_school_all)

df_school_all %>% count(total_visits)
```


# Investigating objects, Base R approach

1. Answer the following questions about the object `df_school_all` by running the appropriate R command within text (as opposed to in an R code chunk) and writing any substantive response required by the question. The first question will be answered for you to show how it works
    + What "type" of object is `df_school_all`?
        + ANSWER [GIVEN]: the object `df_school_all` has type equals `r typeof(df_school_all)`
        type
    + What is the "length" of the object `df_school_all`? What does this specific value of length refer to? (Hint: Use `r ` to run code within text)
        + ANSWER 
    + how many "rows" are in the object `df_school_all`? what does each row represent?  
        + ANSWER 
        
        
\newline  

2. In the below code chunk, use the `str()` function to describe the contents of `df_school_all` and then answer the following questions in text above the code chunk.  
    + What does each element of the object `df_school_all` represent?  
    
    + Are the individual elements within `df_school_all` lists or vectors?  
    
    + Are the individual elements within `df_school_all` named or un-named? what do these element names refer to?  
    
    
```{r}

```

\newline

3. These questions refer to the variable `school_type` within the object `df_school_all`. For the first two questions, run the appropriate R command within text (as opposed to in an R code chunk) and write any substantive response required by the question
    + What is the data "type" of `school_type`?
        + ANSWER
    + What is the "length" of `school_type`? What does this specific value of length refer to?
        + ANSWER
        
4. In these questions, you will apply the `table()` function to the variable `school_type` within the object `df_school_all`.  
    + In your own words, what does the `table()` function do?  
        + ANSWER
    + What does the `useNA` argument of the `table()` function control?
        + ANSWER
    + What is the default value of the `useNA` argument and what does this default value mean?
        + ANSWER
    + What happens when you assign the value `"ifany"` to the `useNA` argument?
        + ANSWER
    + What happens when you assign the value `"always"` to the `useNA` argument?
        + ANSWER
    + In the below R code chunk, use the `table()` function to count the number of observations for each value `school_type` three different ways: 
        + First, without specifying any value for `useNA`
        + Second, by assigning the value `"ifany"` to the `useNA` argument
        + Third,  by assigning the value `"always"` to the `useNA` argument
```{r}

```

# Filter and select questions

The data frame `df_school_all` has one observation for each high school (public and private). 

- The variables that begin with `visits_by_...` identify how many off-campus recruiting visits the high school received from a particular public university. For example, UC Berkeley has the ID `110635` so the variable `visits_by_110635` identifies how many visits the high school received from UC Berkeley. 
- The variable `total_visits` identifies the number of visits the high school received from all (16) public research universities in this data collection sample.


- Before conducting analyses, we'll rename the variable `avgmedian_inc_2564` to give it a shorter name.

1. Rename the variable `avgmedian_inc_2564` to `med_inc` and assign new variable name to the existing object `df_school_all`
```{r}

```



2. The news outlet is interested in comparing the in-state and out-of-state high school visits for [The University of Alabama](https://nces.ed.gov/globallocator/col_info_popup.asp?ID=100751) (IPEDS ID = 100751) variable `visits_by_100751`. Count the number of in-state public high school visits for The University of Alabama.   
    - Note: You will need to use `filter` and the `count` function  
    - Use commas to separate varibles for this questions (e.g. filter(dataframe, variable == something, variable == something))
    - You can do this in one step by wrapping the `count` function around the `filter` function ; or you can do this in two steps by creating a new data frame first  
```{r}

```




3. How many public in-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?
```{r}

```



4. Now count the out-of-state public high school visits for the University of Alabama without using commas.  
    - hint: `&` or `%in%`
```{r}

```




5. How many public out-of-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?
```{r}

```



Once finished, knit to (HTML) and upload both .Rmd and HTML files to class website under the [week 2 tab](https://ozanj.github.io/rclass/resources/)  
*Remember to use this naming convention "lastname_firstname_ps2"*





